DeepPicar: A Low-cost Deep Neural Network-based Autonomous Car

### 2018-01-26 paper/evaluation.tex
Update evaluation section

In this section, we experimentally analyze various real-time aspects
of the DeepPicar. This includes
(1) measurement based worst-case execution time (WCET) analysis of
deep learning inferencing,
(2) the effect of using multiple cores in accelerating inferencing,
(3) the effect of co-scheduling multiple deep neural network models,
and 
(4) the effect of co-scheduling memory bandwidth intensive co-runners.

@@

The Raspberry Pi 3 Model B platform used in DeepPicar equips a Boardcom
BCM2837 SoC, which has a quad-core ARM Cortex-A53 cluster,
running at up to 1.2GHz. Each core has 16K private I\&D caches, and all
cores share a 512KB L2 cache.
The chip also includes Broadcom's Videocore IV
GPU, although we did not use the GPU in our evaluation due to the lack
of sofware support (TensorFlow is not compatible with the Raspberry Pi's GPU).
For software, we use Ubuntu MATE 16.04, TensorFlow 1.1 and Python
2.7. We disabled DVFS (dynamic voltage frequency scaling) and
configured the clock speed of each core statically at the maximum 1.2GHz.

@@

\begin{table}[t]
  \centering
  \begin{tabular} {| c | r | r | r | r |}
    \hline
    \textbf{Operation} & \textbf{Mean} & \textbf{Max} &   \textbf{99pct.} & \textbf{Stdev.} \\ \hline
    Image capture        & 2.28  &  4.94 &  4.54  & 0.52 \\ \hline
    Image pre-processing & 3.09  &  4.60 &  3.31  & 0.10 \\ \hline
    DNN inferencing      & 37.30 & 51.03 & 45.48  & 2.75 \\ \hline
    Total Time           & 42.67 & 56.37 & 50.70  & 2.80 \\ \hline
  \end{tabular}
  \caption{Control loop timing breakdown.}
  \label{tbl:control-loop-breakdown}
\end{table}

@@

First, as expected, we find that the inference operation
dominates the control loop execution time, accounting for about 85\% of
the execution time.
Second, and more importantly, we also find that the measured average
execution time of a single control loop is 42.67 ms, or 23.4 Hz and
the 99 percentile time is 50.70 ms.
This means that the DeepPicar can operate
at about a 20 Hz control frequency in real-time using only the on-board
Raspberry Pi 3 computing platform, as no remote computing resources 
were necessary. We consider these results respectable given the complexity
of the deep neural network, and the fact that the inference operation
performed by TensorFlow only utilizes the CPU cores of the
Raspberry Pi 3 (its GPU is not supported by Tensorflow).

In comparison, NVIDIA's DAVE-2 system, which has the exact same neural
network architecture, reportedly runs at 30 Hz~\cite{Bojarski2016}, which 
is just a bit faster than the DeepPicar. Although we believe it was not
limited by their computing platform (we will experimentally compare
performance differences among multiple embedded computing platforms,
including NVIDIA's Jetson TX2, later in
Section~\ref{sec:comparison}), the fact that the low-cost
Raspberry Pi 3 can achieve similar real-time control performance is
surprising.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 61.96 ms on a single core to 42.67 ms on four
cores (a 30\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 2 cores to 3 cores, the
improvement is mere 2.38 ms (or 4\%). In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
TensorFlow's multicore implementation or if it's the model's inherent
characteristics.

The poor scalability opens up the possibility of consolidating
multiple different tasks or different neural network models rather
than allocating all cores for a single neural network model. For
example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately and
executed on a single dedicated core. Assuming we use the same network
architecture for all models, then one might expect to achieve up to
15 Hz using one core (given 1 core can deliver 62 ms average
execution time). In the next experiment, we investigate the
feasibility of such a scenario.

@@

Figure~\ref{fig:perf-vs-modelcnt} shows the results. In the figure, the
X-axis shows the system configuration: \#of DNN models x \#of CPU
cores/DNN. For example, '4Nx1C' means running four DNN models each of
which is assigned to run on one CPU core, whereas '2Nx2C' means running
two DNN models, each of which is assigned to run on two CPU
cores. The Y-axis shows the average inference timing.
The two bars on the left show the impact of co-scheduling four DNN
models. Compared to executing a single DNN model on one CPU core
(1Nx1C), when four DNN models are co-scheduled (4Nx1C), each model
suffers an average inference time increase of approximately 15 ms,
$\sim$24\%. On the other hand, when two DNN models, each using two CPU
cores, are co-scheduled (2Nx2C), the average inference timing is increased by
about 7 ms, or 10\%, compared to the baseline of running one model
using two CPU cores (1Nx2C).

@@

These increases in inference times in the co-scheduled scenarios are
expected and likely caused by contention in the shared hardware
resources, such as the shared L2 cache and/or the DRAM controller.
To further analyze the source of contention, we use hardware
performance counters of the processor. Specifically, we measure L2
miss rates of the DNN models first in isolation and then after
co-scheduling other models. If the shared L2 cache is the primary
source of inteference, then the measured L2 miss rates will
increase. Figure~\ref{fig:l2missrate-vs-modelcnt} shows the results.
As can be see in the figure, L2 miss rates are largely unchanged
regardless of whether multiple models are co-scheduled or not. This
suggests that the shared L2 cache is not the bottleneck that caused
execution time increases. In other words, DNN models don't appear to be
sensitive to the shared L2 cache space. % L2 partitioning is probably
                                % not going to be useful.
Instead, we hypothesize that it is likely caused by the memory
controller---the only other major shared hardware source---where
memory requests from different CPU cores contend, which would result
in increased memory access latency. While some Intel processors
provide incore hardware counters that can measure average memory
access latency~\cite{ye2016maracas}, we were not able to identify
whether such hardware counters exist in the BCM2837 processor of
Raspberry Pi 3 due to the lack of documentation. Instead, in the next
experiment, we use memory intensive synthetic benchmarks to test the
hypothesis.

@@

We run a single DNN model one core, and
co-schedule an increasing number of the memory intensive synthetic
benchmarks~\footnote{We use the \emph{Bandwidth} benchmark in the
  IsolBench suite, with the following command line parameters: \texttt{\$
  bandwdith -a write -m 16384}}, on the remaining idle cores.

@@

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 9.6X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates do not increase as much. This suggests that
the DNN model's exeuction increase cannot be fully explained by
increases in L2 cache-misses. Instead, as we hypothesized in the previous
experiment, the increased memory pressure from the co-scheduled memory
intensive benchmarks is likely the primary cause of the DNN model's execution
time increase. Therefore, we conclude that DeepPicar's DNN model is
more senstive to DRAM access latency than L2 cache space.

@@

In order to determine if the DNN is mostly affected by shared DRAM 
contention, we partition the L2 cache of the Pi3 and reperform the 
same experiments to see if there are any noticable changes. For the 
partition, we employ PALLOC~\cite{yun2014rtas}, a color-based page 
allocator that works on the kernel level. For the bit mask, we select 
bits 12, 13 and 14 as they can be used to access the L2 cache,
as can be seen by \ref{fig:cache-mapping}. This results in 
2\textsuperscript{3} = 8 colors, which we then assign to the Pi3's 
physical cores such that each core has two unique colors (colors 0 
and 1 are assigned to core 0, colors 2 and 3 are assigned to core 1, 
etc.). In the case of bit 12, since we use 2 colors for each partition,
the L1D cache is not partitioned in our experiments.

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.7\textwidth]{figs/palloc_multicore}
  \caption{Timing impact of co-scheduling multiple DNNs when cache 
partitioning is enabled.}
  \label{fig:palloc_multimodel}
\end{figure}

@@

By partitioning the shared L2 cache, we find that no noticable 
improvements are gained and that performance remains consistent, or 
decreases, depending on the circumstances. As a result, we conclude that 
cache partitioning is not an effective isolation mechanism, and that 
the performance of the shared DRAM is of greater importance to the 
real-time efficiacy of the DeepPicar.

\begin{table*}[!h]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    Item    & Raspberry Pi 3 (B)   & Intel UP                  & NVIDIA Jetson TX2\\
    \hline
            & BCM2837              & X5-Z8350 (Cherry Trail)   & Tegra X2 \\
    CPU     & 4x Cortex-A53@1.2GHz/512KB L2  &
              4x Atom@1.92GHz/2MB L2 &
              4x Cortex-A57@2GHz/2MB L2 \\
            &              &              & 2x Denver@2.0GHz/2MB L2 (not used)  \\
    \hline
    GPU     &  VideoCore IV (not used)    &
               Intel HD 400 Graphics (not used) &
               Pascal 256 CUDA cores   \\
    \hline
    Memory  & 1GB LPDDR2   &  2GB DDR3L     & 8GB LPDDR4              \\
    \hline
	Peak Memory Bandwidth & 8.5 GB/s & 12.8 GB/s & 59.7 GB/s \\
	\hline
	Measured Bandwidth & 2127.94 MB/s & 3951.94 MB/s & 4447.90 MB/s \\
	\hline
  \end{tabular}
  \caption{Compared embedded computing platforms}
  \label{tbl:platforms}
\end{table*}

@@

First, we find that DeepPicar's computing platform,
the Raspberry Pi 3 Model B, offers adquate computing capacity to
perform real-time control of the RC car at 20 Hz frequency (or
50ms per control loop). Given the complexity of the DNN used, we
were pleasantly suprised by this finding. The time breakdown shows that
the DNN inferencing operation, performed by the Tensorflow library,
dominates the execution time, which is expected.

Second, we find that scalability of Tensorflow's DNN 
implementation is limited. We find that using all four cores is 
only about 30\% better than using just a single core.

>>>>

In this section, we experimentally analyze various real-time aspects
of the DeepPicar. This includes
(1) measurement based worst-case execution time (WCET) analysis of
deep learning inferencing,
(2) the effect of using multiple cores in accelerating inferencing,
(3) the effect of co-scheduling multiple deep neural network models,
and 
(4) the effect of co-scheduling memory bandwidth intensive co-runners.
(5) the effect of partitioning the shared L2 cache.

@@

The Raspberry Pi 3 Model B platform used in DeepPicar equips a Boardcom
BCM2837 SoC, which has a quad-core ARM Cortex-A53 cluster,
running at up to 1.2GHz. Each core has a 16K private Instruction cache
and a 32K private Data cache, and all cores share a 512KB L2 cache.
The chip also includes Broadcom's Videocore IV
GPU, although we did not use the GPU in our evaluation due to the lack
of sofware support (TensorFlow is not compatible with the Raspberry Pi's GPU).
For software, we use Ubuntu MATE 16.04, TensorFlow 1.1 and Python
2.7. We disabled DVFS (dynamic voltage frequency scaling) and
configured the clock speed of each core statically at the maximum 1.2GHz.

@@

\begin{table}[t]
  \centering
  \begin{tabular} {| c | r | r | r | r |}
    \hline
    \textbf{Operation} & \textbf{Mean} & \textbf{Max} &   \textbf{99pct.} & \textbf{Stdev.} \\ \hline
    Image capture        & 1.61  &  1.81 &  1.75  & 0.05 \\ \hline
    Image pre-processing & 2.77  &  2.90 &  2.87  & 0.04 \\ \hline
    DNN inferencing      & 18.49 & 19.30 & 18.99  & 0.20 \\ \hline
    Total Time           & 22.87 & 23.74 & 23.38  & 0.20 \\ \hline
  \end{tabular}
  \caption{Control loop timing breakdown.}
  \label{tbl:control-loop-breakdown}
\end{table}

@@

First, as expected, we find that the inference operation
dominates the control loop execution time, accounting for about 81\% of
the execution time.
Second, and more importantly, we also find that the measured average
execution time of a single control loop is 22.87 ms, or 43.7 Hz and
the 99 percentile time is 23.38 ms.
This means that the DeepPicar can operate
at a 30 Hz control frequency in real-time using only the on-board
Raspberry Pi 3 computing platform, as no remote computing resources 
were necessary. We consider these results surprising given the complexity
of the deep neural network, and the fact that the inference operation
performed by TensorFlow only utilizes the CPU cores of the
Raspberry Pi 3 (its GPU is not supported by Tensorflow).

In comparison, NVIDIA's DAVE-2 system, which has the exact same neural
network architecture, reportedly runs also at 30 Hz~\cite{Bojarski2016}. 
Although we believe it was not
limited by their computing platform (we will experimentally compare
performance differences among multiple embedded computing platforms,
including NVIDIA's Jetson TX2, later in
Section~\ref{sec:comparison}), the fact that the low-cost
Raspberry Pi 3 can achieve similar real-time control performance is
surprising.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 46.30 ms on a single core to 22.86 ms on four
cores (over a 100\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 3 cores to 4 cores, the
improvement is mere 2.80 ms, or 12%. In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
TensorFlow's multicore implementation or if it's the model's inherent
characteristics.

The poor scalability opens up the possibility of consolidating
multiple different tasks or different neural network models rather
than allocating all cores for a single neural network model. For
example, it is conceivable to use four cameras and four different
neural network models, each of which is trained separately and
executed on a single dedicated core. Assuming we use the same network
architecture for all models, then one might expect to achieve up to
20 Hz using one core (given 1 core can deliver 46 ms average
execution time). In the next experiment, we investigate the
feasibility of such a scenario.

@@

Figure~\ref{fig:perf-vs-modelcnt} shows the results. In the figure, the
X-axis shows the system configuration: \#of DNN models x \#of CPU
cores/DNN. For example, '4Nx1C' means running four DNN models each of
which is assigned to run on one CPU core, whereas '2Nx2C' means running
two DNN models, each of which is assigned to run on two CPU
cores. The Y-axis shows the average inference timing.
The two bars on the left show the impact of co-scheduling four DNN
models. Compared to executing a single DNN model on one CPU core
(1Nx1C), when four DNN models are co-scheduled (4Nx1C), each model
suffers an average inference time increase of approximately 11 ms,
$\sim$24\%. On the other hand, when two DNN models, each using two CPU
cores, are co-scheduled (2Nx2C), the average inference timing is increased by
about 4 ms, or 13\%, compared to the baseline of running one model
using two CPU cores (1Nx2C).

@@

These increases in inference times in the co-scheduled scenarios are
expected and likely caused by contention in the shared hardware
resources, such as the shared L2 cache and/or the DRAM controller.
To further analyze the source of contention, we use hardware
performance counters of the processor. Specifically, we measure L2
miss rates of the DNN models first in isolation and then after
co-scheduling other models. If the shared L2 cache is the primary
source of inteference, then the measured L2 miss rates will
increase. Figure~\ref{fig:l2missrate-vs-modelcnt} shows the results.
As can be see in the figure, L2 miss rates increase as more models are 
co-scheduled, but the increase is less than expected.
This suggests that the shared L2 cache is not the main bottleneck that 
causes execution time increases. In other words, DNN models don't appear 
to be significantly affected by the shared L2 cache space. 
% L2 partitioning is probably not going to be useful.
Instead, we hypothesize that it is likely caused by the memory
controller---the only other major shared hardware source---where
memory requests from different CPU cores contend, which would result
in increased memory access latency. While some Intel processors
provide incore hardware counters that can measure average memory
access latency~\cite{ye2016maracas}, we were not able to identify
whether such hardware counters exist in the BCM2837 processor of
Raspberry Pi 3 due to the lack of documentation. Instead, in the next
experiment, we use memory intensive synthetic benchmarks to test the
hypothesis.

@@

We run a single DNN model on one core, and
co-schedule an increasing number of the memory intensive synthetic
benchmarks~\footnote{We use the \emph{Bandwidth} benchmark in the
  IsolBench suite, with the following command line parameters: \texttt{\$
  bandwdith -a write -m 16384}}, on the remaining idle cores.

@@

Figure~\ref{fig:} shows the normalized execution time and L2 miss-rate
of the DNN model running on the Core 0 as a function of the number of
co-scheduled memory intensive synthetic benchmarks. First, as we
increase the number of co-runners, the DNN model's execution times are
increased---by up to 11.6X---even though the DNN model is running on a
dedicated core (Core 0). On the other hand, the DNN model's L2
cache-miss rates do not increase as much. This suggests that
the DNN model's execution increase cannot be fully explained by
increases in L2 cache-misses. Instead, as we hypothesized in the previous
experiment, the increased memory pressure from the co-scheduled memory
intensive benchmarks is likely the primary cause of the DNN model's execution
time increase. Therefore, we conclude that DeepPicar's DNN model is
more senstive to DRAM access latency than L2 cache space.

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.7\textwidth]{figs/palloc_multimodel}
  \caption{Timing impact of co-scheduling multiple DNNs when cache 
partitioning is enabled.}
  \label{fig:palloc_multimodel}
\end{figure}

In order to determine if the DNN is mostly affected by the DRAM 
controller, we partition the L2 cache of the Pi3 and reperform the 
same experiments to see if there are any noticable changes. For the 
partition, we employ PALLOC~\cite{yun2014rtas}, a color-based page 
allocator that works on the kernel level. For the bit mask, we select 
bits 12, 13 and 14 as they can be used to access the L2 cache,
as can be seen by Figure \ref{fig:cache-mapping}. This results in 
2\textsuperscript{3} = 8 colors, which we then assign to the Pi3's 
physical cores such that each core has two unique colors (colors 0 
and 1 are assigned to core 0, colors 2 and 3 are assigned to core 1, 
etc.). In the case of bit 12, since we use 2 colors for each partition,
the L1 Data cache is not partitioned in our experiments.

@@

@@

By partitioning the shared L2 cache, we find that no noticable 
improvements are gained and that performance remains consistent, or 
decreases, depending on the circumstances. As a result, we conclude that 
cache partitioning is not an effective isolation mechanism, and that 
the performance of the shared DRAM controller is of greater importance 
to the real-time efficiacy of the DeepPicar.

@@

First, we find that DeepPicar's computing platform,
the Raspberry Pi 3 Model B, offers computing capacity to
perform real-time control of the RC car at 30 Hz frequency (or
33.$\overline{\mbox{33}}$ ms  per control loop). Given the complexity of 
the DNN used, we were pleasantly surprised by this finding. 
The time breakdown shows that the DNN inferencing operation, 
performed by the Tensorflow library, dominates the execution time, 
which is expected.

Second, we find that scalability of Tensorflow's DNN 
implementation is limited. We find that using all four cores can double 
performance of the DNN, but compared to using three cores, performance
improvement is relatively minimal.
